### 2022-11-16 chapter1.tex
Fixed using bad quotation marks

\subsection*{Initial alignment and selection of representative sequences}
Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes. As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph. For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a "bridge" formed by the genomes which contain only one of the isoforms. If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences. However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences. Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states” \cite{Felsenstein1996}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by "noise" from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocess roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the "gap stickiness" as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.

@@

\subsection*{Inference of species trees}
Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences. In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition. Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences. However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}. Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups. We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A). To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were "reverse translated" from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B). Both trees have an identical topology, which is consistent with other published phylogenies \cite{D12GC2007, DaLage2007}.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{trees/out/trees1.png}
\centering
\caption{\textbf{Phylogenetic tree of species.}
\textbf{(A)} Consensus tree from LG model fit to meta-alignments directly sampled from the original protein alignments. \textbf{(B)} Consensus tree from GTR model fit to meta-alignments sampled from "reverse translated" nucleotide alignments. Values at nodes are bootstrap percentages.}
\label{fig:trees}
\end{figure}

\section*{Discussion}
% TODO: Highlight specific contributions
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus. To assist these efforts, the final alignments after curation and the labels from the "missing" phylo-HMM are provided as supplemental data. Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well. While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated. However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work. Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.

>>>>

\subsection*{Initial alignment and selection of representative sequences}
Since the NCBI annotation pipeline incorporates transcriptome data from a variety of sources, its inputs are heterogeneous in sequencing depth, developmental stage, and tissue of origin across different genomes. As a result, some genomes are annotated with different or multiple splice isoforms of a given orthologous gene, which can create complex networks in the resulting hit graph. For example, if the genomes are variably annotated with one or both of two distinct isoforms, the resulting graph may contain two clusters connected by a ``bridge'' formed by the genomes which contain only one of the isoforms. If the nodes bridging the two clusters form cliques with themselves and the clusters, the clique percolation algorithm will merge all the nodes into a single orthologous group where some genes have multiple associated sequences. However, these additional sequences can complicate downstream comparative analyses that may not easily generalize to genes with multiple associated sequences. Thus, in our pipeline a single representative was chosen for each gene using an alignment-based strategy detailed in the methods section.

@@

Phylogenetic HMMs (phylo-HMMs) are statistical models that incorporate phylogenetic and spatial information to calculate the probability that each observation in a sequence was generated by one of several “hidden states” \cite{Felsenstein1996}. Since they can evaluate both the probability of a gap pattern in a column given the known phylogenetic relationships and the local context, a phylo-HMM was used to segment the alignment into contiguous regions with different patterns of gaps. A fully specified phylo-HMM requires a fixed number of hidden states and a probability distribution for each. Thus, we identified four distinct types of regions in the alignments, roughly corresponding to highly conserved regions with few to no gaps, divergent regions, regions with a stable gap pattern discordant with the expected phylogenetic relationships, and regions with poorly supported segments. For simplicity, however, we refer to the states that generate each type of region as 1A, 1B, 2, and 3, respectively. To model probability distributions for each state, we first conceptualized the observed alignments as the superposition of two distinct processes (Fig.~\ref{fig:hmm}A, left). The first is a phylogenetic process which evolves and splits a single ancestral sequence over time according to a tree. The second is the annotation process which can erroneously exclude or include segments from a sequence. The result is an alignment of annotated sequences which contains evolutionary information obscured by ``noise'' from the annotation process, shown here by the exclusion of three N-terminal residues in the fourth annotated sequence. To simplify modeling this behavior with an HMM, we coded the sequences into binary symbols. The distributions for each state then consisted of two components derived from the encoded sequences (Fig.~\ref{fig:hmm}A, right). The first component models the gap pattern with a Markov process. This Markov process is in turn composed of two subprocesses where the first is a phylogenetic process, and the second is a jump process. These subprocess roughly correspond to changes caused by evolution and annotation, respectively. Because this first component did not fully capture the propensity for the gap patterns to remain constant, we included a second component that models the ``gap stickiness'' as a beta-binomial random variable by counting the number of symbols that remain constant between columns. Each component is associated with a set of parameters, and the unique parameters for each state yield its characteristic gap pattern and gap stickiness.

@@

\subsection*{Inference of species trees}
Many phylogenetic methods require a species tree to inform the evolutionary relationships between sequences. In fact, the phylo-HMMs discussed previously used a species tree as an input, though we omitted this detail for clarity of exposition. Therefore, to support the curation step and other downstream analyses, we sought to infer phylogenetic trees from the aligned sequences. However, since the roots of phylogenetic trees are not identifiable with commonly used time-reversible substitution models, we repeated the orthology inference pipeline with the outgroup species \textit{Scaptodrosophila lebanonensis}. Afterwards, we inferred phylogenetic trees using the LG model of amino acid substitution from 100 meta-alignments sampled from alignments of single copy orthologous groups. We then combined them into a single consensus tree (Fig.~\ref{fig:trees}A). To provide a similar tree for the analysis of non-coding regions, we inferred phylogenetic trees using the GTR model of nucleotide substitution from 100 meta-alignments sampled from nucleotide alignments which were ``reverse translated'' from the protein alignments and their corresponding coding sequences (\ref{fig:trees}B). Both trees have an identical topology, which is consistent with other published phylogenies \cite{D12GC2007, DaLage2007}.

\begin{figure}[h!]
\includegraphics[width=\textwidth]{trees/out/trees1.png}
\centering
\caption{\textbf{Phylogenetic tree of species.}
\textbf{(A)} Consensus tree from LG model fit to meta-alignments directly sampled from the original protein alignments. \textbf{(B)} Consensus tree from GTR model fit to meta-alignments sampled from ``reverse translated'' nucleotide alignments. Values at nodes are bootstrap percentages.}
\label{fig:trees}
\end{figure}

\section*{Discussion}
% TODO: Highlight specific contributions
The orthologous groups and alignments yielded by this pipeline are a valuable resource for comparative studies of gene birth/death processes and protein evolution at the level of both entire proteomes and specific gene families in the \textit{Drosophila} genus. To assist these efforts, the final alignments after curation and the labels from the ``missing'' phylo-HMM are provided as supplemental data. Although this work focused on single copy orthologs, other studies may require different subsets of orthologous groups that demand other pre-processing and alignment strategies. Therefore, we have included the orthologous groups and the initial alignments with and without non-representative sequences in the supplemental data as well. While we anticipate these resources will remain relevant in the near term, the trends that permitted this work to substantially improve on previous efforts will render them obsolete in the coming years as more \textit{Drosophila} genomes are assembled and annotated. However, an authoritative and lasting set of orthologous groups and alignments is not the primary goal of this work. Instead, it serves as a case study in how dense taxonomic sampling and modern genome assembly and annotation pipelines present new opportunities and challenges to the traditional techniques for identifying and aligning orthologous groups.
